Governing through patronage: the bargain for education in decentralized Brazil

### 2020-09-11 paper/paper.tex
finalize edits for paper to brazil data lab

\begin{abstract}
    Why do politicians engage in patronage and what are its consequences for public service delivery? I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators. Leveraging fieldwork evidence and administrative data, I demonstrate that mayors with a weaker hold on the local legislature resort to patronage to co-opt legislators. Educational staff turnover has negative downstream effects on student learning, but limited electoral backlash. Weak electoral accountability channels suggests that patronage is primarily a political elite game, with little involvement by the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

@@

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations. How can these educational outcomes be improved? Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo? Who holds power over public education, and whose interests is it meant to serve?

To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. But the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In it, the exchange of public sector appointments for political favors -- a canonical example of patronage -- was designed to grant the mayor legislative support to govern according to her policy agenda.

@@

The estimation proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct ffect on patronage. Building on a canonical model, I propose a theoretical model of legislative vote-buying from which I derive my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of specifications and different measures of turnover.

@@

The paper is structured as follows. Section 1 provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. In section 2 I present the main theoretical argument, with a formal treatment of the subject. Section 3 presents the institutional context and data, followed in section 4 by the research design and main results. Section 5 concludes.

\section{Related literature}

@@

\section{Institutional context and data}

@@

Additional data has been collected to supplement the estimation exercise. Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\url{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}, (2)\url{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}, (3)\url{https://www.seduc.ce.gov.br/spaece/}.}

@@

\section{Empirical results}

@@

Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in \citet{imai_covariate_2014} which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

@@

Finally, I present some evidence of differential returns to patronage, in particular for city councilors. For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education.

@@

\section{Conclusion}

@@

% ## Covariate balance

% ### Education quality vs. staff turnover

>>>>

\begin{abstract}
    How can public education be improved? I argue that local politicians govern through patronage, undermining education at the expense of voters. Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites. Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector. The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain. Patronage induces turnover in educational staff, with negative downstream effects on student learning. Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

@@

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations. How can public education be improved? Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors relate to one another in local institutions? Who holds power over public education, and whose interests is it meant to serve?

To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

@@

The estimation proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct effect on patronage. Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of specifications and different measures of turnover.

@@

The paper is structured as follows. Section \ref{sec:literature} provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. Section \ref{sec:context} presents the context and data. In section \ref{sec:theory}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model. Section \ref{sec:empirics} presents the research design and main results. Section 5 concludes.

\section{Related literature}
\label{sec:literature}

@@

\section{Institutional context and data}
\label{sec:context}

@@

Additional data has been collected to supplement the estimation exercise. Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\hyperlink{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}{IBGE}, (2)\hyperlink{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}{Finbra}, (3)\hyperlink{https://www.seduc.ce.gov.br/spaece/}{Spaece}.}

@@

\section{Empirical results}
\label{sec:empirics}

@@

% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in \citet{imai_covariate_2014} which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

@@

Finally, I present some evidence of differential returns to patronage, in particular for city councilors. For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education. The negative consequences of patronage on student learning, however, do not bite, as nayors who deliver a worse quality of education are no less likely to be reelected in the next term.\footnote{See appendix \ref{app:accountability}.}

@@

\section{Conclusion}
\label{sec:conclusion}

@@

\subsection{Covariate balance}

\begin{figure}[h]
    \centering
    \includegraphics{plots/covariate_balance.pdf}
\end{figure}

\subsection{Accountability of mayors}
\label{app:accountability}

\begin{figure}[h]
    \centering
    \includegraphics{plots/accountability_fit.pdf}
\end{figure}

\subsection{Regression Discontinuity Design}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\linewidth]{plots/visual_rdd_legislative_seat_flip.pdf}
\end{figure}
